SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2019-09-08 scipy-1.0/paper.tex
Streamlined SciPy begins section

At this point, scientific Python started attracting more serious attention;
code that started out as side projects by graduate students had grown into
essential infrastructure at national laboratories and research institutes.
For example, Paul Dubois at Lawrence Livermore National Laboratory (LLNL) took over the
maintenance of Numeric and funded the writing of its
manual\cite{Numeric-manual}, and
the Space Telescope Science Institute (STScI), which was in charge of
Hubble Space Telescope science operations, decided to replace their
custom scripting language and analysis pipeline with Python\cite{STScI-slither}.

As SciPy, the algorithms library of the ecosystem, began attracting the attention
of large research organizations,
the next generation of graduate students and postdocs were already exploring
other aspects of the computational environment.
In 2000, Prabhu Ramachandran, a PhD student at the Indian Institute of
Technology, started work on a 3D visualization application, based on
Kitware's C++ Visualization Toolkit\cite{schroeder:2006:VTK}, called
Mayavi\cite{mayavi-intro}.
In 2001, Fernando Pérez, a graduate student at the University of
Colorado, Boulder, created the IPython interactive shell, which
eventually blossomed into Project Jupyter\cite{Kluyver:2016aa}.
% IPython hosted on SciPy:
% https://mail.python.org/pipermail/scipy-user/2003-March/001455.html
John Hunter, a postdoc
%John got his PhD in 2001, then a post-doc with John Milton (dept of
%Neurology, UChicago?) a couple of years; then research professor
%under Kurt Hecox, whereafter at Tradelink from 2006
at the University
of Chicago, liked the plotting functionality
available in MATLAB\cite{matlab}, but had problems accessing their laboratory's
license, which was governed by a hardware dongle.  In response, he
wrote a plotting library from scratch, and Matplotlib 0.1 was released
April 2003\cite{matplotlib-rel}.

% https://mail.python.org/pipermail/scipy-dev/2002-June/001007.html
% https://str.llnl.gov/str/News896.html

%2002—https://web.archive.org/web/20021020215937/http://www.scipy.org:80/site_content/scipy02/scipy02_agenda.htm
%2003—https://web.archive.org/web/20031207151811/http://www.scipy.org:80/site_content/scipy03/
%2004—Jim Hugunin
%https://web.archive.org/web/20040920080225/http://www.scipy.org:80/wikis/scipy04
%Travis Oliphant 2005—https://web.archive.org/web/20051103155042/http://www.scipy.org:80/wikis/scipy05/FrontPage
%Guido—2006 https://scipy.github.io/old-wiki/pages/SciPy2006.html
%Martelli—2008 https://scipy.github.io/old-wiki/pages/Developer_Zone/Conferences/SciPy2004.html
%Norvig—2009 https://scipy.github.io/old-wiki/pages/Developer_Zone/Conferences/SciPy2005.html

As STScI continued to use Python for an increasingly large portion
of the Hubble Space Telescope data analysis pipeline, they encountered
problems with the Python numerical array container.
Numeric, the original array package, was
suitable for small arrays, but not for the large images processed by
STScI\@.  With the Numeric maintainer's blessing, the decision was made
to write NumArray\cite{greenfield2003numarray}, a library that could handle data on a larger
scale.  Unfortunately, NumArray proved inefficient for small arrays,
presenting the community with a rather unfortunate choice.  In 2005,
Travis Oliphant combined the best elements of Numeric and NumArray,
thereby solving the dilemma and paving the way for Python to become
a very significant player in the Data Science movement. NumPy
1.0 was released in October 2006\cite{numpy-1.0-tag}, with about 30
authors recognized for major contributions in its release notes.

% Future directions for SciPy after meeting at Berkeley
% https://mail.python.org/pipermail/scipy-user/2005-March/004164.html

%In April 2006, Paul
%Dubois\footnote{https://mail.python.org/pipermail/numpy-discussion/2006-April/007487.html} wrote
%to the NumPy mailing list:
%\begin{quote}
%IEEE's magazine, Computing in Science and Engineering (CiSE), has asked me
%to put together a theme issue on the use of Python in Science and
%Engineering. I will write an overview to be accompanied by 3-5 articles [. . .]
%that show a diverse set of applications or
%tools, to give our readers a sense of whether or not Pyth%on might be useful
%in their own work.
%\end{quote}
In a May/June 2007 special issue of IEEE Computing in Science and
Engineering, Paul Dubois wrote\cite{dubois2007guest}:
\begin{quote}
LLNL now has many Python-based efforts built from scratch or wrapped around
legacy codes \textelp{} hundreds of thousands
of lines of C++, Python, and Fortran~95, all working together just as we hoped,
doing compute-intensive calculations on massively parallel computers.
\end{quote}
While it is now more commonly accepted, Dubois took the time to explain to
the 2007 reader that ``interpreted doesn't mean slow or only interactive.''
He also shared his own interest in Python for ``computational steering''
where ``Python serves as the input language to a scientific application, and the actual
computations are performed both in Python itself and in compiled extensions.''
%https://www.computer.org/csdl/mags/cs/2007/03/c3007.html
Using Python for computational steering was one of the earliest uses of Python
in large scientific processing pipelines.
In addition to Dubois' overview there were also articles introducing
Python and SciPy \cite{oliphant2007python}, IPython \cite{perez2007ipython},
and  Matplotlib \cite{hunter2007matplotlib}.
There was also a diverse set of
research applications including
systems biology\cite{myers2007python},
astronomy\cite{greenfield2007reaching},
robotics\cite{krauss2007python},
nanophotonics\cite{bienstman2007python},
partial differential equations\cite{mardal2007using},
neuroimaging\cite{millman2007analysis},
geographic information systems\cite{shi2007python}, and
education\cite{backer2007computational, myers2007pythona}.

>>>>

At this point, scientific Python started attracting more serious attention;
code that started out as side projects by graduate students had grown into
essential infrastructure at national laboratories and research institutes.
For example, Paul Dubois at Lawrence Livermore National Laboratory (LLNL) took over the
maintenance of Numeric and funded the writing of its
manual\cite{Numeric-manual}, and
the Space Telescope Science Institute (STScI), which was in charge of
Hubble Space Telescope science operations, decided to replace their
custom scripting language and analysis pipeline with Python\cite{STScI-slither}.
As STScI continued to use Python for an increasingly large portion
of the Hubble Space Telescope data analysis pipeline, they encountered
problems with the Python numerical array container.
Numeric, the original array package, was
suitable for small arrays, but not for the large images processed by
STScI\@.  With the Numeric maintainer's blessing, the decision was made
to write NumArray\cite{greenfield2003numarray}, a library that could handle data on a larger
scale.  Unfortunately, NumArray proved inefficient for small arrays,
presenting the community with a rather unfortunate choice.  In 2005,
Travis Oliphant combined the best elements of Numeric and NumArray,
thereby solving the dilemma. NumPy 1.0 was released in October
2006\cite{numpy-1.0-tag}, paving the way for the reunified scientific
Python community to mature.
